\plaintitle

### 2016-05-24 paper/paper.tex
Some changes paper

\usepackage[
    backend=biber,
    style=draft
]{biblatex}

@@

\begin{abstract}
    Visualization of sports data is very common in recent years as access to
    data has increased. The abundance of data and audience allows for new and
    innovative presentations. In this paper we describe an interactive
    visualization of the NBA competition over the last 30 years. With three
    different views, we allow the user to explore data about NBA seasons
    including standings, team statistics, transfers and player statistics. The
    d3.js framework was used to develop this visualization. 
\end{abstract}

@@

The National Basketball Association (NBA) is the most famous professional
basketball league in the world\cite{nbawiki}. As in every other sports
branch, a lot of data and statistics about players, teams, games and seasons
are available, mainly in large tables. Despite the large amount of data
available, gathering useful insights from this data can be surprisingly
challenging. In our work we try to present this data in an innovative and
intuitive way, allowing the user to draw meaningful conclusions.

In section~\ref{sec:goal} we describe the goal of the visualization and the
target audience. In section~\ref{sec:data} we describe the data used, including
its origins, advantages and limitations. In section~\ref{sec:literature} we give
an overview of related literature and web resources, including related
visualizations.  This includes both visualizations of NBA or other sports data,
as well as visualizations tackling an isue encountered during the development of
our visualization. In section~\ref{sec:visualization} we describe the
visualization itself. We give an overview of the different stages of development
of the visualization, as well as the major design decisions.
Section~\ref{sec:discussion} discusses potential improvements of the final
visualization and lessons learned from the project. We conclude in
section~\ref{sec:conclusion}. Appendix~\ref{sec:terminology} contains
definitions for common basketball terms used in this paper.

\section{Goal and Target Audience}\label{sec:goal} 
The visualization's goal can be best summarized by the following sentence:
``Why did a certain NBA-team win the Championship?'' The visualization focuses
on exposing relations between team performance and their player roster over
several years. It allows users to find explanations for major improvements and
declines in team performance.  The visualization allows exploration of NBA data
by lay persons.  More specifically, the visualization does not focus on premade
explanations for phenomenons visible in the NBA data. By providing easy and
intuitive access to the data, the visualization allows users to draw their own
conclusions. The visualization's target audience are lay persons. Specifically,
basketball fans are the core of our target audience. It can be of particular
interest with the recent surge in fantasy leagues, for which data analysis is
paramount to a player's success~\cite{fantasy,fantasyskill}.

\section{Data}\label{sec:data}
The data visualized is a subset of the data available on basketball statistics
site basketball-reference\cite{basketball-reference}. A wide range of data is
available on this site. This includes common basketball performance statistics,
such as field goals, percentage of shots scored, shooting distance, minutes
played per season, number of personal fouls made, salaries and much more. All of
this data is available as time-series. These
statistics are avalable for each player individually for each season. Aggregated
data for entire teams and for entire seasons is also available. Additionally,
individual game statistics are also available, including playoff games.  In our
visualization we only use data from 1984 onwards. This is done because of rule
changes in the NBA which changed the number of teams competing and the
competition's structure. To simplify implementation of our visualization, only
data after the last major rule change was used.  The data we use includes league
standings and playoff rankings for each team, team overall statistics, the
team's roster and individual player statistics for each year, including the
PER~(Player Efficiency Rating)\cite{per}.

@@

\begin{figure}
\centering
  \includegraphics[width=1.0\columnwidth]{figures/data}
  \caption{An overview of the data.}
  \label{fig:data}
\end{figure}

\section{Related Work}\label{sec:literature}
Recently, there has been an increase in the interest in visualizations of sports
data. This is best exemplified by theworkshop on sports data visualization at
the IEEE Visualization Conference in 2013\cite{ieeevis}.

@@

Tan et al. created an adaption of the traditional tournament bracket to better
support moden interaction\cite{adaptivitree}. The tool, AdaptiviTree, changes
the brackets shape and adds colored lines to show available information as the
user pciks his own bracket.

@@

The entire history of the NBA and the best teams were analyzed by
fivethirtyeight~\cite{fivethirtyeight}. They calculated an elo score for all
teams an visualized them on line graphs.

@@

This technique follows the standards of information visualization. It provides
an overview, allows zooming and filtering and provides details on
demand\cite{mantra,multipleviews,automatingdesign}. The play-off view provides
the full overview which allows filtering by team. The statistics view provides
the ability to zoom into one specific team. There some details are provided
instantly, while others can be procured by hovering over certain elements.

The organization of these three views is in a vertical layout with fixed
scrolling. Navigation with the keyboard is supported as well.

The choice was made to use a thematic background instead of opting for the
traditional minimalistic approach to information visualization. This was done to
improve the attractiveness of the visualisation on the one hand and the users'
memory of it on the other hand\cite{aesthetics}.

All three views are discussed in more detail below. The three views are
available to the user on a single webpage. The user start on the play-off view
and is able to navigate to the statistics and team views by scrolling down or
clicking on teams. A different season can be selected with the left and right
arrows or by clicking on the timeline. The timeline is a fixed part of the
visualization. It is visible in the three different views.

\subsection{The play-off view}
This view informs the user of the roster and end positions in the NBA play-offs
for a selected season, in general or for a selected team. The view consists of
two parts: 
\begin{itemize}
    \item Context section: the top part informs the user about the selected team
    and shows a legend of the play-off view.
    \item Bubble section: this section represents the play-offs of the selected
        season.
\end{itemize}
The play-off view is shown in figure~\ref{fig:playoffviewwithcontext}.

\subsubsection{Context section}
The context section contains a legend for the bubble section. The bubble's
borders are gold, silver or bronze if the team ended respectively first, second
or third in the regular competition. Teams that didn't end in a podium place
during the regular competition have a red or blue stroke color indicating the
team's region, respectively the western or eastern conference. The context
section also contains a link to an ``about'' page where the user can get an
explanation of the visualization with its different views, can get some
background information about the NBA and the terminolgy used, and can check the
source of our used data. If the user hovers over or selects a team in the bubble
section, the context section shows more information about that team, as shown in
Figure~\ref{fig:playoffviewteamselected}. The information shown includes the
Simple Rating System score (SRS), the end position in the play-offs and the end
position in the regular competition.

@@

The curved lines between teams represent that those teams have encountered each
other in the playoffs. Each row represents an end position in the play-offs. The
higher a team is, the further it has come.  From the top to the bottom the rows
represent the winner of the final (rank 1), loser of the final (rank 2), losers
of the semi-finals (rank 3), losers of the quarter-finals (rank 4), losers of
the eighth finals (rank 5). The last row represents the teams that did not make
it to the play-offs (rank 6).

This simple presentation intuitively puts the winning teams higher than the
losing teams. The winner of a certain game can be easily deriving from their
relative vertical position. It is also clear that each team neads to have played
and won a series of matches against one other team at every level below
them~\ref{fig:playoffviewteamselected}. This view is further reinforced by
highlighting a team and its matches and opponents when hovering over it.
Additional team information is then also shown in the context part.

@@

\subsubsection{Previous designs}
Our initial design used the team circles' lightness to indicate their final
ranking in their region. The hue was used to show the league they were part of.
According to the Ranking of Perceptual Tasks~\cite{perceptualranking} this is a
good property for ordered values like the regular competition ranking. But the
combination of the logo's of the teams with the colors resulted in an overload
of informaiton and made it more difficult to compare the end ranking of
different teams.

As an alternative we tried the same view without the logo's. This was deemed
inconvenient because you cannot instantly see which team is represented by which
circle. Therefore we've chosen to only use the teams' logos.

Our initial play-off view was too dense because of the connections between teams
as well. It connected teams that played against each other directly. When
searching for alternatives to reflect on our choice we discovered a similar
visualization in a totally different context\cite{whitehousepath}.  This
visualization gives a cleaner view of the competitors for a specific team
compared to our original sketch. Hence we decided to adapt our visualization. We
added an intermediate step between two teams to more clearly indicate how teams
competed to become NBA champion.

@@

\subsection{The statistics view}
The statistics view gives a user a clear overview of how a statistic has been
influenced in the selected season compared to the season before. The statistics
view is showed in Figure~\ref{fig:statisticsview}. The view consists of 3
coordinated parts:
\begin{itemize}
    \item Context section: the top part informs the user of the context of the
        selected team
    \item Arrow section: this section illustrates why a statistic changed
        compared to the previous season
    \item Selected statistics section: team statistics and how they evolve over
        time
\end{itemize}

@@

\subsubsection{Arrow section of selected statistic}
This section illustrates a change for a chosen statistic for the selected team
compared to the previous season. This gives a user more insight into why a
statistic changed over time and how this change has impacted team performance.
The selected team is represented by a circle in the same way as explained in the
play-off view. The change for a statistic is represented by three arrows:
\begin{itemize}
    \item Arrow on the left pointing towards the circle: indicating the
        influence of players who joined the team 
    \item Arrow on the right of the circle: indicating how the team internally
        changed, e.g. by training its players.
    \item Arrow on the left pointing away from the circle: indicating the
        influence of players who left the team 
\end{itemize}
\todo{add statistics view image here}
The jerseys next to the arrows are the jerseys of players that had the biggest
contribution to that arrow. On the left, different statistics , including player
PER can be chosen from. Below this, player information is shown when hovering
over a player's jersey.

\subsubsection{Selected statistics}
Below the arrow section some selected statistics are shown over
time.  The line charts enables a user to identify peaks or drops that
influenced the outcome of a team is this season, following seasons or previous
seasons. A user interacts with the visualization by shifting the bar indicating
the current year displayed. By doing this the view gets updated and the user can
see the impact on the team in the context section and why the statistic changes
in the arrow section. This view provides a lot of additional information to
users by allowing them to see correlations between statistics and phenomena
quickly and easily.

\subsection{The team view}
This view shows each player of the team on its field positions. Players are
represented by their jerseys. Jerseys aligned next to each other share the same
position, which is indicated as well. The size of a jersey would illustrate a
players performance on this spot. This view as well as the statistics view
should give the user the ability to search for explanations why a team is is
performing better or worse over seasons. On the other hand, a user can see what
impact a change in team characteristics has on its overall performance. This
view should give a user insight on how strong or weak a particular team is for a
specific field position. A user can then scroll through time to see how field
positions evolve and what the influence is over time.

\subsubsection{Previous designs}
Initially we designed this view more elaborate. When the user would click on a
jersey, specific player information would be shown. A number of player
statistics would be shown as small multiples. A bell curve should have informed
how a player scores compared to his team, other players in the league on this
position or all other players in the league. We considered other charts for the
small multiples as well. A box plot or bullet was less clear for us to indicate
where a player stands compared to the group he was compared with. Alternatives
for the small multiples were evaluated as well. This part of the visualization
was deemed less important than the others and has not been fully developed
because of time and resource constraints. \todo{sketch of small multiples
here}

\subsection{Improving the result}\todo{Is this necessary? I don't know what to
write here}
Talk about how we improved the data-ink ration in the end..

\section{Future work}
As with most projects, the result is never 'finished', meaning that there are
always improvements possible but due to timing or resource constraints they
haven't all been implemented. We suggest a number of features that might improve
the visualization:
\begin{itemize}
    \item Smooth transitions between states could improve the user experience.
        For example when changing year, a transformation of the arrow in the
        statistics view would help the user see a difference between years.  At
        the moment the arrows change at the blink of an eye, making it difficult
        for us humans with a short term memory to make the comparison. This
        would have been especially useful in the transitions between different
        years on the bubble view.
    \item To further support our short term memory we could also 'save' the
        previous state. Eg. in the field view we could visualize ghost jerseys
        to indicate which players left the team or position and highlight
        players that joined the team.  This would help the user see how the team
        changes over time.
    \item At the moment no attention has been payed to the process, provenance
        and history of the user's exploration session. A user thus needs to
        memorize his action to be able to share or reproduce his actions. This
        could be improved by adding URL parameters. This way a user at least can
        share or save a state of an analysis. One step further is to provide a
        play-feature. The visitor could them share his exploration and replay
        it. A simple implemention would ask the visitor for a start date and an
        end date and would then 'play' the changes of a particular team and
        characteristic over time.
    \item We could support users further in looking for patterns by focussing
        more on comparisons. We could integrate a view where a visitor is able
        to compare two teams in the same year, or in two different years.
\end{itemize}

\subsection{Technology}
To create the visualization, the d3 javascript framework~\cite{d3} was used in
combination with html5~\cite{html5} and jquery~\cite{jquery}.  Note that D3 is
based on HTML, CSS, Javascript and SVG. This allows easy access to the
visualization as most modern browsers are capable of handling these
webstandards.  The choice not to use the d3 framework for the entire
visualization was made to ease the layout configuration of the visualization.
Instead, html was used to do the global layout of the visualization. We opted to
structure the site with multiple divs as containers. To each container we
allocate a svg with a specific visualization.

\todo{sketch of how the page is structured here}
When working with d3.js we encountered some obstacles:
\begin{itemize}
        \item A synchrounous call need to be combined with synchronous calls.
            More specific, when a visitor uses the timeline to scroll through
            time, the whiping of previous visualizations was not synced with the
            creation of the visualisation. This resulted in whiping sections
            when a visualization was not created yet. Hence the view resulted in
            multiple figures overlapping.
\end{itemize}

\section{Lessons learned}\label{sec:discussion}
We've learned a couple of lessons during the creation of this visualization:
\begin{itemize}
    \item Find a story to tell with your visualization
    \item Each separate component should be a sufficient informative visualization on itself
    \item Creating a custom visualization costs time and opportunities
\end{itemize}
In what follows we will discuss our experience in more detail.

@@

\subsection{Each component should be informative}
In order to have a good visualization as a whole, each component should be a
clear and informative visualization on itself. During the project, we started noticing that most
components in a visualization lack references and hence are not informative
enough. Eg. each small multiple of a statisctic in the statistics view on itself
gives sufficient information to stand on itselfs. The bar in the small multiple
indicates the score in the selected season and enables the user to evaluate that
score over time. Have there been better scores, or worse scores. Is this score
part of an upward movement during years? This should be weighed against the
duplication of data and use of screen space. Coordinates views alleviate this
need significantly as well, as each component is strongly supported by other
components.

\subsection{Custom visualizations}
Before exploring what is out there in the d3.js world, we made our own sketches.
This is how we came up with the 3 custom views that make up our visualization.
When comparing our design with alternatives we could not find something that
satisfied our needs. Nonetheless, they gave us inspiration to finetune our
designs. Although we are satisfied with our result and we are convinced it was
the right choice to reach our goal, it had some draw backs. We noted that other
teams had more freedom in exploring multiple different visualizations which
enabled them to evaluate different options with trial and error. Much of the
time developing was spent on the technical details for these custom
visualizations. This was especially true for the bubble component of the
playoffs view. No out of the box solution for this could be found so everything
from the determination of the positions of the bubbles to how to connect them
had to be developed by us.

@@

% REFERENCES FORMAT
% References must be the same font size as other body text.
%\bibliographystyle{SIGCHI-Reference-Format}
\printbibliography
%TODO References should be in alphabetical order and represented by numbers in the text.

>>>>

\usepackage[backend=biber, style=draft]{biblatex}
%\usepackage{custom}

@@

\begin{abstract}
\todo{Abstracts should be about 150 words}
    Visualization of sports data is very common in recent years as access to
    data has increased. The abundance of data and audience allows for new and
    innovative presentations. In this paper we describe an interactive
    visualization of the NBA competition over the last 30 years. With three
    different views, we allow the user to explore data about NBA seasons
    including standings, team statistics, transfers and player statistics. The
    d3.js framework was used to develop this visualization. 
\end{abstract}

@@

The National Basketball Association (NBA) is the most famous professional
basketball league in the world\cite{nbawiki}. As in every other sports
branch, a lot of data and statistics about players, teams, games and seasons
are available, mainly in large tables. Despite the large amount of data
available, gathering useful insights from this data can be surprisingly
challenging. In our work we try to present this data in an innovative but still
intuitive way, allowing the user to draw meaningful conclusions.

\todo{reference to sections should use the name and not a number because the numbers are not shown, so they are not usable}
In section~\ref{sec:goal} we describe the goal of the visualization and the
target audience. In section~\ref{sec:data} we describe the data used, including
its origins, advantages and limitations. In section~\ref{sec:literature} we give
an overview of related literature and web resources, including related
visualizations.  This includes both visualizations of NBA or other sports data,
as well as visualizations tackling an issue encountered during the development of
our visualization. In section~\ref{sec:visualization} we describe the
visualization itself. We give an overview of the different stages of development
of the visualization, as well as the major design decisions.
Section~\ref{sec:discussion} discusses potential improvements of the final
visualization and lessons learned from the project. We conclude in
section~\ref{sec:conclusion}. Appendix~\ref{sec:terminology} contains
definitions for common basketball terms used in this paper.

\section{Goal and Target Audience}\label{sec:goal} 
The visualization's goal can be best summarized by the following sentence:
``Why did a certain team win the NBA Championship?'' The visualization focuses
on exposing relations between team performance and their player roster over
several years. It allows users to find explanations for major improvements and
declines in team performance.  The visualization allows exploration of NBA data
by lay persons.  More specifically, the visualization does not focus on premade
explanations for phenomenons visible in the NBA data. By providing easy and
intuitive access to the data, the visualization allows users to draw their own
conclusions. The visualization's target audience are lay persons. Specifically,
basketball fans are the core of our target audience. It can be of particular
interest with the recent surge in fantasy leagues, for which data analysis is
paramount to a player's success~\cite{fantasy,fantasyskill}.

\section{Data}\label{sec:data}
The data visualized is a subset of the data available on basketball statistics
site basketball-reference\cite{basketball-reference}. It is a wide range of data including common basketball performance statistics, such as field goals, percentage of shots scored, shooting distance, minutes played per season, number of personal fouls made, salaries and much more . All of this data is available as time-series. These statistics are available for each player individually for each season. Aggregated data for entire teams and for entire seasons is also available. Additionally, individual game statistics are also available, including playoff games. In our visualization we only use data from 1984 onwards. This is done because of rule changes in the NBA which changed the number of teams competing and the
competition's structure. To simplify implementation of our visualization, only
data after the last major rule change was used.  The data we use includes league
standings and playoff rankings for each team, team overall statistics, the
team's roster and individual player statistics for each year, including the
PER~(Player Efficiency Rating)\cite{per}.

@@

\todo{Figure is still in dutch}
\begin{figure}
\centering
  \includegraphics[width=1.0\columnwidth]{figures/data}
  \caption{An overview of the data.}
  \label{fig:data}
\end{figure}

\section{Related Work}\label{sec:literature}
Recently, there has been an increase in the interest in visualizations of sports
data. This is best exemplified by the workshop on sports data visualization at
the IEEE Visualization Conference in 2013\cite{ieeevis}.

@@

Tan et al. created an adaption of the traditional tournament bracket to better
support moden interaction\cite{adaptivitree}. The tool, AdaptiviTree, changes
the brackets shape and adds colored lines to show available information as the
user picks his own bracket.

@@

The entire history of the NBA and the best teams were analyzed by
fivethirtyeight~\cite{fivethirtyeight}. They calculated an ELO-score \todo{explanation of the elo score?} for all
teams an visualized them on line graphs.

@@

The technique used, follows the standards of information visualization. It provides
an overview, allows zooming and filtering and provides details on
demand\cite{mantra,multipleviews,automatingdesign}. The play-off view provides
the full overview which allows filtering by team. The statistics view provides
the ability to zoom into one specific team. There some details are provided
instantly, while others can be procured by hovering over certain elements.

All three views are discussed in more detail below. The three views are
available to the user on a single webpage, where they are organized in a 
vertical layout with fixed scrolling. Navigation with the keyboard is 
supported as well. The user start on the play-off view and is able to 
navigate to the statistics and team views by scrolling down or clicking 
on teams. A different season can be selected with the left and right
arrows or by clicking on the timeline. The timeline is a fixed part of the
visualization. It is visible in the three different views. Navigation 
with the keyboard for switching between the three different views is 
supported as well.

We did not opt for the traditional minimalistic approach used in information 
visualizations. The data-ink design principles were kept in mind, but we also 
focused on aesthetics instead of maximizing the data-ink ratio. So the choice 
was made to use a thematic background. This was done to improve the 
attractiveness of the visualisation on the one hand and the users' memory of 
it on the other hand\cite{aesthetics}.

\subsection{The play-off view}
This view informs the user of the roster and end positions in the NBA play-offs
for a selected season, in general or for a selected team. The play-off view is 
shown in figure~\ref{fig:playoffviewwithcontext}.The view consists of
two parts: 
\begin{itemize}
    \item Context section: the top part informs the user about the selected team
    and shows a legend of the play-off view.
    \item Bubble section: this section represents the play-offs of the selected
    season.
\end{itemize}

\subsubsection{Context section}
The context section contains a legend for the bubble section. The bubble's
borders are gold, silver or bronze if the team ended respectively first, second
or third in the regular competition. Teams that didn't end on a podium place
during the regular competition have a red or blue stroke color indicating the
team's region, respectively the western or eastern conference. The context
section also contains a link to the ``About'' page where the user can get an
explanation of the visualization with its different views, can get some
background information about the NBA and the terminolgy used, and can check the
source of the data used in the visualization. If the user hovers over or selects 
a team in the bubble section, the context section shows more information about that 
team, as shown in Figure~\ref{fig:playoffviewteamselected}. The information shown 
includes the Simple Rating System score (SRS), the end position in the play-offs 
and the end position in the regular competition.

@@

The curved lines between teams represent the fact that those teams have 
encountered each other in the playoffs. Each row represents an end position 
in the play-offs. The higher a team is, the further it has come.  From the 
top to the bottom the rows represent the winner of the final (rank 1), loser 
of the final (rank 2), losers of the semi-finals (rank 3), losers of the 
quarter-finals (rank 4), losers of the eighth finals (rank 5). The last row 
represents the teams that did not make it to the play-offs (rank 6).

This simple presentation intuitively puts the winning teams higher than the
losing teams. The winner and loser of a certain game can be easily derived from 
their relative vertical position. It is also clear that for a team to be on a 
higher level, the team needs to have played and won a series of matches against 
one other team at every level below them~\ref{fig:playoffviewteamselected}. 
This view is further reinforced by highlighting a team and its matches and 
opponents when hovering over it. Additional team information is then also shown 
in the context part.

@@

\subsubsection{Previous designs}
Our initial design used the team circles' lightness to indicate their final
ranking in their region. The hue was used to show the league they were part of.
According to the Ranking of Perceptual Tasks~\cite{perceptualranking} this is a
good property for ordered values like the regular competition ranking. But the
combination of the logo's of the teams with the colors resulted in an overload
of information and made it more difficult to compare the end ranking of
different teams.

As an alternative we tried the same view without the logo's. It turned out this 
was inconvenient because you could not instantly see which team is represented 
by which circle. Therefore we've chosen to only use the teams' logos.

Our initial play-off view was too dense because of the connections between teams
as well. Teams that played against each other were connected directly. When
searching for alternatives to reflect on our choice we discovered a similar
visualization in a totally different context\cite{whitehousepath} as explained in 
the section \ref{sec:literature}. With the idea of this visualization in mind, our 
visualization gives a cleaner view of the competitors for a specific team compared 
to our original sketch. Hence we decided to adapt our visualization. We added an 
intermediate step between two teams to more clearly indicate how teams competed to 
become the NBA champion. \todo{meaning of last sentence is not clear}

@@

\subsection{The statistics view}
The statistics view gives a user a clear overview of how a statistic has been
influenced in the selected season compared to the season before. The statistics
view is showed in Figure~\ref{fig:statisticsview}. The view consists of 3
coordinated parts:
\begin{itemize}
    \item Context section: the top part informs the user of the context of the
        selected team
    \item Arrow section: this section illustrates why a statistic has changed
        compared to the previous season
    \item Selected statistics section: team statistics and how they evolve over
        time
\end{itemize}

@@

\subsubsection{Arrow section}
This section illustrates a change for a chosen statistic for the selected team in 
the selected season compared to the previous season. This gives a user more 
insight into why a statistic changed over time and how this change has affected 
team performance. The selected team is represented by a circle in the same way as 
explained in the play-off view. To give the user a better interpretation of the SRS 
value of the circle, three other reference circles are drawn. These are 
transparant with a lightgrey border. They show the maximum, average and minimum SRS 
value of the NBA in the selected season. The change for the chosen statistic is 
represented by three arrows:
\begin{itemize}
    \item Arrow on the left pointing towards the circle: indicating the
        influence of players who joined the team.
    \item Arrow on the right of the circle: indicating how the team internally
        changed, e.g. by training its players
    \item Arrow on the left pointing away from the circle: indicating the
        influence of players who left the team 
\end{itemize}
\todo{add statistics view image here}
The value of the change in statistic is shown next to the arrow. The number of 
players responsible for the change is shown between brackets. Their are shown 
jerseys next to the arrows as well. These are the jerseys of the two players that 
had the biggest contribution to the influence represented by that arrow.  
On the left, there is a possibility to chose a statistic. The possibilities 
are the PER value, the age, the effective field goal percentage, the offensive rating 
and the defensive rating. Below this, player information is shown when a user hovers
over a player's jersey.

\subsubsection{Selected statistics section}
Below the arrow section, multiple line and area charts show how the selected 
team has performed over the years. The different statistics shown are the 
total PER of the whole team, the league rank, the playoff rank, the average age 
of the team, the audience, the SRS-value of the team, the offensive rating and 
the defensive rating of the team. A user interacts with the visualization by 
shifting the bar indicating the current year displayed or the timeline. By 
doing this the view gets updated and the user can see the impact on the team 
in the context section and why the statistic changes in the arrow section. 
These charts enable the user to identify peaks or drops in a certain statistic. 
By interacting with the visualization, previous seasons can be inspected to try 
to find explanations or following seasons can be inspected to see if a team 
responds in a specific way or what the consequences are. This view provides 
a lot of additional information to users by allowing them to see correlations 
between statistics and phenomena quickly and easily.

\subsubsection{Previous designs}
This view has not changed much compared to the initial design. The only difference 
was the position of the arrow and jerseys for the players who stayed, which were 
originally in the team circle, and so the logo was not showed. 
The design of the arrows did not change. The incoming and outgoing 
arrows are maybe more intuitevly than the arrow for the stayed players, but we did 
not find any alternatives we could use.

\subsection{The team view}
The team view gives a user a detailed view on the strength of the team on the 
different field positions. The team view is showed in Figure~\ref{fig:teamview}. 
The view consists of 3 coordinated parts:
\begin{itemize}
    \item Context section: the top part informs the user of the context of the
        selected team
    \item Field section: this section illustrates why a statistic has changed
        compared to the previous season
    \item Selected statistics section: team statistics and how they evolve over
        time
\end{itemize}
The context section and the selected statistics section are fixed parts in the 
statistics and team view, and are therefore not explained again for this view.

\subsubsection{Field section}
This view shows each player of the team on its field positions. Players are
represented by their jerseys. Jerseys aligned next to each other share the same
position, which is indicated as well. The size of a jersey illustrates a
players performance according to the chosen statistic. The possible statistics 
are shown on the left and are the same as those in the arrow section of the 
statistics view. As well as the statistics view, this view should give the user 
the ability to search for explanations why a team is performing better or worse 
over seasons. On the other hand, a user can see what impact a change in team 
characteristics has on its overall performance. This view should give a user 
insight on how strong or weak a particular team is for a specific field position. 
A user can then scroll through time to see how field positions evolve and what 
the influence is over time.

\subsubsection{Previous designs}
Initially we designed this view more elaborate. When the user would click on a
jersey, specific player information would be shown together with a number of player
statistics in the form of small multiples. A bell curve should have informed
how a player scores compared to his team, other players in the league on his
position or all other players in the league. We considered other charts for the
small multiples as well. A box plot or bullet was less clear for us to indicate
where a player stands compared to the group he was compared with. This part of the 
visualization was not our main focus. Because of the lesser importnace and because of time and resource constraints, it has not been fully developed. \todo{sketch of small multiples
here}

\subsection{Improving the result}\todo{Is this necessary? I don't know what to
write here.}
Talk about how we improved the data-ink ration in the end..

\section{Future work}
As with most projects, the result is never 'finished', meaning that there are
always extensions and iprovements possible but due to timing or resource constraints they haven't all been implemented. We suggest a number of features that might improve the visualization:
\begin{itemize}
    \item Smooth transitions between states could improve the user experience.
        For example when changing year, a transformation of the arrow in the
        statistics view would help the user see a difference between years.  At
        the moment the arrows change at the blink of an eye, making it difficult
        for us humans with a short term memory to make the comparison. This
        would also have been very useful in the transitions between different
        years on the bubble view.
    \item To further support our short term memory we could also 'save' the
        previous state. Eg. in the field view we could visualize ghost jerseys
        to indicate which players left the team or position and highlight
        players that joined the team.  This would help the user see how the team
        changes over time.
    \item At the moment no attention has been payed to the process, provenance
        and history of the user's exploration session. A user thus needs to
        memorize his action to be able to share or reproduce his actions. This
        could be improved by adding URL parameters. This way a user at least can
        share or save a state of an analysis. One step further is to provide a
        play-feature. The visitor could them share his exploration and replay
        it. A simple implemention would ask the visitor for a start date and an
        end date and would then 'play' the changes of a particular team and
        characteristic over time.
    \item We could support users further in looking for patterns by focussing
        more on comparisons. We could integrate a view where a visitor is able
        to compare two teams in the same year, or in two different years.
\end{itemize}

\subsection{Technology}
To create the visualization, the d3 javascript framework~\cite{d3} was used in
combination with html5~\cite{html5} and jquery~\cite{jquery}. Note that D3 is
based on HTML, CSS, Javascript and SVG. This allows easy access to the
visualization as most modern browsers are capable of handling these
webstandards. The choice not to use the d3 framework for the entire
visualization was made to ease the layout configuration of the visualization.
Instead, html was used to do the global layout of the visualization. We opted to
structure the site with multiple divs as containers. To each container we
allocate a svg with a specific visualization.
\todo{sketch of how the page is structured here}

When working with d3.js we encountered one main obstacle:
\begin{itemize}
        \item A synchrounous call need to be combined with synchronous calls.
            More specific, when a visitor uses the timeline to scroll through
            time, the whiping of previous visualizations was not synced with the
            creation of the visualisation. This resulted in whiping sections
            when a visualization was not created yet. Hence the view resulted in
            multiple figures overlapping.
\end{itemize}

\section{Lessons learned}\label{sec:discussion}
We've learned a couple of things during the creation of this visualization:
\begin{itemize}
    \item Find a story to tell with your visualization
    \item Each separate component should be a sufficient informative visualization on itself
    \item Creating a custom visualization costs time and opportunities
\end{itemize}
In what follows we will discuss our experience in more detail.

@@

\subsection{Each component should be informative}
In order to have a good visualization as a whole, each component should be a
clear and informative visualization on itself. During the project, we started noticing that most components in a visualization lack references and hence are not informative enough. Eg. each small multiple of a statisctic in the statistics view on itself gives sufficient information to stand on itselfs. The bar in the small multiple indicates the score in the selected season and enables the user to evaluate that score over time. Have there been better scores, or worse scores. Is this score
part of an upward movement during years? This should be weighed against the
duplication of data and use of screen space. Coordinated views alleviate this
need significantly as well, as each component is strongly supported by other
components.

\subsection{Custom visualizations}
Before exploring what is out there in the d3.js world, we made our own sketches.
This is how we came up with the 3 custom views that make up our visualization.
When comparing our design with alternatives we could not find something that
satisfied our needs. Nonetheless, they gave us inspiration to finetune our
designs. Although we are satisfied with our result and we are convinced it was
the right choice to reach our goal, it had some draw backs. If we would have not started with our custom view, we would have had more freedom in exploring multiple different visualizations and we would be able to evaluate different options with trial and error. Instead much of the time developing was spent on the technical details for these custom visualizations. This was especially true for the bubble component of the play-off view. No out of the box solution for this could be found so everything from the determination of the positions of the bubbles to how to connect them had to be developed by us.

@@

% REFERENCES FORMAT
% References must be the same font size as other body text.
%\bibliographystyle{SIGCHI-Reference-Format}

\printbibliography
\todo{References should be in alphabetical order and represented by numbers in the text. Style should be set to style{SIGCHI-Reference-Format}, this moment it is set to the style draft}
